---
title: "Numerical Analysis (data activities)"
author: "Choyee Wolff"
date: "2025-02-23"
output: html_document
---

```{r}
library(haven)
Data_Service <- read_sav("C:/Users/kylew/OneDrive/Desktop/uni/Numerical/8011spss_115789D2C668C0532E10DF66A3148DDB_V1/Data.Service.sav")
View(Data_Service)
```

## Data Activity 1 
### A quick statistic summary for the Data Service

```{r}
summary(Data_Service)
```

### summary only for antisocx variable
```{r}
summary(Data_Service$antisocx)
```
###### Boxplot

```{r}
boxplot(Data_Service$antisocx, main = "Boxplot of antisocx", ylab = "Values", col = "lightblue")
```
###### Many outliers are present on the upper end. It has a positive skew. 

## Data Activity 2

```{r}
names(Data_Service)
```

```{r}
library(forcats)
library(haven)
Data_Service$bcsvictim <- as.character(Data_Service$bcsvictim)  # Convert to character
Data_Service$bcsvictim <- as.factor(Data_Service$bcsvictim)  # Convert to factor

table(Data_Service$bcsvictim)
```

## Data Activity 3
Create a subset of individuals who belong to the ‘75+’ age group and who were a ‘victim of crime’ that occurred in the previous 12 months. Save this dataset under a new name 'crime_75victim'.

```{r}
library(dplyr)

# Check the class of agegrp7
class(Data_Service$agegrp7)

# Check the class of bcsvictim
class(Data_Service$bcsvictim)
```
Convert the varables to characters 
```{r}
Data_Service$agegrp7 <- as.character(as_factor(Data_Service$agegrp7))

# Convert bcsvictim to a character (since it's a factor, this ensures correct filtering)
Data_Service$bcsvictim <- as.character(Data_Service$bcsvictim)
```

Filter

```{r}
crime_75victim <- Data_Service %>%
  filter(agegrp7 == "75+" & bcsvictim == "1")

# Check the first few rows
head(crime_75victim)
```

## Data Activity 4

Create a boxplot for the variable 'antisocx'
If you’re using ‘graphics’: Add “Levels of anti-social behaviour in neighbourhood ‘antisocx’” as a title and colour the plot in purple and colour the outliers in blue.

```{r}
boxplot(Data_Service$antisocx, 
        main = "Levels of anti-social behaviour in neighbourhood ‘antisocx’", 
        col = "purple", 
        outcol = "blue",  # Outliers in blue
        ylab = "antisocx Levels")
```
```{r}
crime_counts <- table(Data_Service$bcsvictim)

# Create a bar plot with an orange colour
barplot(crime_counts, 
        main = "Survey Respondents Experiencing Crime in the Last 12 Months", 
        col = "orange",
        xlab = "Crime Victim (0 = No, 1 = Yes)", 
        ylab = "Frequency")
```

 Bar plot of bcsvictim in orange to show the number of crime victims
 
 
 
## Data Activity 5
Upload new data 
Find out mean, median and mode of variables sbp, dbp and income.


```{r}
library(haven)
Health_Data <- read_sav("C:/Users/kylew/Downloads/Health Data.sav")
```

```{r}
# Mean
mean(Health_Data$sbp, na.rm = TRUE)
mean(Health_Data$dbp, na.rm = TRUE)
mean(Health_Data$income, na.rm = TRUE)

# Median
median(Health_Data$sbp, na.rm = TRUE)
median(Health_Data$dbp, na.rm = TRUE)
median(Health_Data$income, na.rm = TRUE)

# Mode function
calc_mode <- function(x) {
  uniq_x <- unique(na.omit(x))  # Remove NA values
  uniq_x[which.max(tabulate(match(x, uniq_x)))]  # Find most frequent value
}

calc_mode(Health_Data$sbp)
calc_mode(Health_Data$dbp)
calc_mode(Health_Data$income)


```
 Five-Number Summary of income and Boxplot
 
```{r}
# Five-number summary of income
summary(Health_Data$income)

boxplot(Health_Data$income, 
        main = "Income Distribution", 
        col = "lightblue", 
        ylab = "Income")
```
 
To check if systolic blood pressure (sbp) is different between people with and without peptic ulcers, we use an independent t-test.

```{r}
# Ensure peptic ulcer is a factor
Health_Data$peptic_ulcer <- as.factor(Health_Data$pepticulcer)

# Perform independent t-test
t_test_result <- t.test(sbp ~ pepticulcer, data = Health_Data, var.equal = TRUE)

# Print results
print(t_test_result)
```
p-value > 0.05 → There is no significant association between sbp and peptic ulcer presence.

### scatter plot

```{r}
library(ggplot2)

# Scatter plot using ggplot2
ggplot(Health_Data, aes(x = income, y = sbp)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot of SBP vs. Income",
       x = "Income",
       y = "Systolic Blood Pressure (SBP)") +
  theme_minimal()
```
Bar plot of bcsvictim in orange to show the number of crime victims

## Data Activity 6

Nonparametric Tests Using Health_Data in R

This activity involves nonparametric statistical tests to analyze age, dbp (diastolic blood pressure), and sbp (systolic blood pressure).

```{r}
# Mean of age
mean(Health_Data$age, na.rm = TRUE)

# Median of age
median(Health_Data$age, na.rm = TRUE)

# Mode function (most frequent value)
calc_mode <- function(x) {
  uniq_x <- unique(na.omit(x))
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}

# Mode of age
calc_mode(Health_Data$age)

```
 Test Whether Median Diastolic BP (dbp) is the Same in Diabetic vs. Non-Diabetic Participants
 
```{r}
# Ensure 'diabetes' is a factor variable
Health_Data$diabetes <- as.factor(Health_Data$diabetes)

# Wilcoxon test for median differences in diastolic BP
wilcox_test_result <- wilcox.test(dbp ~ diabetes, data = Health_Data)

# Print result
print(wilcox_test_result)

```
 p-value > 0.05 → No significant difference in dbp between groups.

Since occupation has multiple categories, we use the Kruskal-Wallis Test (nonparametric ANOVA alternative).

```{r}
# Ensure 'occupation' is a factor variable
Health_Data$occupation <- as.factor(Health_Data$occupation)

# Kruskal-Wallis test for systolic BP across occupation groups
kruskal_test_result <- kruskal.test(sbp ~ occupation, data = Health_Data)

# Print result
print(kruskal_test_result)

```
p-value > 0.05 → No significant difference in sbp between occupations.

## Data activity 7 
```{r}
library(dplyr)
crime_table <- table(Data_Service$bcsvictim, Data_Service$agegrp7)
crime_table
prop.table(table(Data_Service$bcsvictim, Data_Service$agegrp7), margin = 1) 
```

prop.table() converts a contingency table into proportions (percentages or probabilities), making it easier to analyse distributions.

25-34 is the most likely victims and 75+ is the least likely victims


